Article Title

### 2014-01-27 plos.tex
split analysis of structure in Results into two sections:

Estimator $C_{\sf sparse}$ is produced by approximating the sample covariance matrix whose inverse is sparse: 
\begin{equation}\label{eq:c-sparse}
C_{\sf sparse} = S^{-1}.
\end{equation}
Here $S$ is a sparse matrix, \ie one in which a large fraction of off-diagonal elements are set to zero.  The estimator has one hyperparameter that determines the sparsity (fraction of off-diagonal zeros) in $S$.  The structure imposed by $C_{\sf sparse}$ favors conditions in which neural correlations arise from direct linear effects (\sq{interactions}) between some pairs of neurons (\figref{02}{Row 1, C}).  We will sometimes refer to sparse partial correlations between neurons simply  as \sq{interactions}.

@@

\paragraph{Simulation}
We next demonstrat how the most efficient among different regularized estimators can reveal the structure of correlations.
We constructed four $50\times 50$ covariance matrices, each with structure that matched one of the four regularized estimators (\figref{02}{\,Row 2, A--D}, see Methods for details about how these matrices were generated).  We used these covariance matrices as the ground truth in multivariate normal distributions with zero means and drew samples of various sizes. Even with relatively large sample sizes (\eg $n=1000$), the sample correlation matrices were contaminated by sampling noise and their underlying structures were difficult to discern (\figref{02}{\,Row 3}).

@@

\paragraph{Covariance estimation in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{03}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{03}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} to produce (\figref{03}{C and D}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was presented for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{03}{E}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with between 100 and 300 repetitions of each direction. In this stimulus, each grating lasted 1 second and was followed by a 1-second blank.  The average stimulus response was subtracted from each trial. The traces were then binned into 150 ms intervals for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low (0.01--0.03 in most images sites) (\figref{03}{F and G}).

@@

In these dense populations, direct interactions between cells are likely to influence the patterns of population activity.  We therefore hypothesized that covariance matrix estimates that explicitly model the partial correlation between pairs of neurons ($C_{\sf sparse}$ and $C_{\sf sparse+latent}$) would have a performance advantage.  However, the observed neurons must also be strongly influenced by global activity fluctuations and by unobserved common inputs, to the advantage of estimators that explicitly model common fluctuations of the entire population: $C_{\sf factor}$ and $C_{\sf sparse+latent}$.  If both types of effects are significant, then $C_{\sf sparse+latent}$ should outperform the other estimators.

@@

Furthermore, $C_{\sf sparse}$ outperformed all other estimators except $C_{\sf sparse+latent}$ (Fig.~S\ref{supp:01}) suggesting the greater importance of modeling interactions between pairs of neurons than the common fluctuations of activity across the entire population.

@@

\paragraph{Correlation structure and circuit architecture}
Because of its superior estimation performance, $C_{\sf sparse+latent}$ likely provides a more physiologically relevant representation of the functional connectivity than the other estimators.  The estimator segregates two sources of correlations:  The first source was the network of linear dependencies between pairs of neurons. The second source was the set of latent units reflecting common fluctuations across the entire population.  These could be caused by  common inputs, cell assembly activation, network fluctuations or other forms of synchronized activations.

We examined the composition of the $C_{\sf sparse+latent}$ estimates at each imaged site. Although the regularized estimates were similar to the sample correlation matrix (\figref{05}{A and D}), the corresponding partial correlation matrices differed substantially (\figref{05}{B and E}). The partial correlation matrix of the regularized estimate was decomposed into a sparse component and a low-rank component (\figref{05}{C}).  While sample correlations were mostly positive, the sparse partial correlations contained a high fraction of negative interactions.  In the example site (Fig.~\ref{fig:03} and Fig.~\ref{fig:05}),  the sparse component had 82.2\% sparsity (or, conversely, 17.8\% connectivity) with average node degree of 52.8 (\figref{05}{G}).  The estimate also inferred 15 latent units contributing to the low-rank component of the partial correlation matrix.

@@

We compared the graphs of interactions inferred by $C_{\sf sparse+latent}$ to the graph obtained by thresholding the sample correlations.   Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013}.  We set the threshold to match the connectivity of the $C_{\sf sparse+latent}$ graph.  The networks revealed by the two methods differed substantially. In the example site with 17.8\% connectivity inferred by $C_{\sf sparse+latent}$, only 41.7\% of connections coincided with above-threshold sample correlations (\figref{05}{F, H, and I}).   In particular, the thresholded sample correlations were predominantly positive whereas 43.3\% of the interactions obtained using $C_{\sf sparse+latent}$ were negative.  Most of the negative interactions corresponded to low sample correlations (\figref{05}{F}) where high correlations were expected given the rest of the correlation matrix.

The average partial correlations revealed by $C_{\sf sparse+latent}$ estimates at all 31 sites were about 5 times lower and less variable across sites than sample correlations (\figref{06}{D}): the coefficient of variation of the average sample correlations across all sites was 0.41 whereas that of the average partial correlations was only 0.35.  The average node degree of the sparse component of the partial correlations and the number of inferred latent units varied widely between sites but generally increased with recorded population size (\figref{06}{A and B}). However, there was an inverse relationship between the number of latent units and the average node degree (\figref{06}{C}). Several sites, even with relatively large population sizes, had fairly few pairwise interactions and were dominated by latent units.  These differences may be explained by differences in brain states and recording quality.  For example neuropil contamination, motion and other factors may have degraded the quality of the data.

@@

We also examined the relationship of the correlation structure to the differences in orientation preference and to the physical distances between the cell pairs (Fig.\;\ref{fig:07}).  The partial correlations of the $C_{\sf sparse+latent}$ estimates fell more rapidly with the difference in preferred orientation (\figref{07}{A}) than sample correlations. They also decreased more rapidly with physical distance, both laterally  (\figref{07}{B}) and in depth (\figref{07}{B}).

\begin{figure}[!ht]
    \begin{center}
        \includegraphics{./figures/Figure07.pdf}
    \end{center}
    \caption{{\bf Dependence of sample correlations and regularized partial correlations on orientation tuning and physical distance.}
    {\bf A--C.} Average partial correlations (red) estimated by $C_{\sf sparse+latent}$ and average sample correlations (black) averaged across multiple imaged sites. At each site the correlations were normalized by the respective average correlation shown in \figref{06}{D}.  The number of sites, $n$, that qualified to be included in the analysis is indicated in each panel. Sites were included if they had at least 20 pairs of neurons in each of the intervals. The error bars indicate the standard error of the mean based on on the number of sites $n$.
    {\bf A.} Average correlations between pairs of neurons tuned to orientation with differences in preferred orientation in the intervals of 0--15$^\circ$, 15--45$^\circ$ and 45--90$^\circ$. 
    {\bf B.} Average correlations between pairs of neurons located at the same depth ($\pm$25$\mu$m) separated by lateral distances in the intervals of 0--25 $\mu$m, 25--75 $\mu$m, 75--150 $\mu$m, and 150+ $\mu$m.
    {\bf C.} Average correlations between pairs of neurons displaced laterally by less than 25 $\mu$m separated in depth by distances in the intervals of 0-25 $\mu$m, 25--60 $\mu$m, and 60+ $\mu$m.
    {\bf D--F.} Normalized connectivity of positive (green) and negative (dark red) interactions from the sparse component obtained from $C_{\sf sparse+latent}$. Normalized connectivity was computed as the fraction of pairs connected by interactions of corresponding signs in each interval divided by the fraction of non-zero interactions across the entire site. }
\label{fig:07}
\end{figure}

Furthermore, the positive and negative interactions in the $C_{\sf sparse+latent}$ estimator were organized differently: The probability of positive interactions fell rapidly with differences in the preferred orientation (\figref{07}{D}), lateral displacements (\figref{07}{E}), and displacement in depth (\figref{07}{F}) while the negative interactions were much less selective (\figref{07}{D--F}).

>>>>

Estimator $C_{\sf sparse}$ is produced by approximating the sample covariance matrix whose inverse is sparse: 
\begin{equation}\label{eq:c-sparse}
C_{\sf sparse} = S^{-1}.
\end{equation}
Here $S$ is a sparse matrix, \ie one in which a large fraction of off-diagonal elements are set to zero.  The estimator has one hyperparameter that determines the sparsity (fraction of off-diagonal zeros) in $S$. The problem of finding the optimal set of non-zero elements of the precision matrix is known as \emph{covariance selection} \cite{Dempster:1972}. The structure imposed by $C_{\sf sparse}$ favors conditions in which neural correlations arise from direct linear effects (\sq{interactions}) between some pairs of neurons (\figref{02}{Row 1, C}).  We will sometimes refer to sparse partial correlations between neurons simply  as \sq{interactions}.

@@

\paragraph{Simulation}
We next demonstrate how the most efficient among different regularized estimators can reveal the structure of correlations.
We constructed four $50\times 50$ covariance matrices, each with structure that matched one of the four regularized estimators (\figref{02}{\,Row 2, A--D}, see Methods for details about how these matrices were generated).  We used these covariance matrices as the ground truth in multivariate normal distributions with zero means and drew samples of various sizes. Even with relatively large sample sizes (\eg $n=1000$), the sample correlation matrices were contaminated by sampling noise and their underlying structures were difficult to discern (\figref{02}{\,Row 3}).

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{03}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{03}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} to produce (\figref{03}{C and D}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was presented for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{03}{E}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with between 100 and 300 repetitions of each direction. In this stimulus, each grating lasted 1 second and was followed by a 1-second blank.  The average stimulus response was subtracted from each trial. The traces were then binned into 150 ms intervals for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low (average across sites of 0.02 and all under 0.04) (\figref{03}{G} and \figref{06}{D}).

@@

In these dense populations, direct interactions between cells are likely to influence the patterns of population activity.  We therefore hypothesized that covariance matrix estimates that explicitly model the partial correlations between pairs of neurons ($C_{\sf sparse}$ and $C_{\sf sparse+latent}$) would have a performance advantage.  However, the observed neurons must also be strongly influenced by global activity fluctuations and by unobserved common inputs, to the advantage of estimators that explicitly model common fluctuations of the entire population: $C_{\sf factor}$ and $C_{\sf sparse+latent}$.  If both types of effects are significant, then $C_{\sf sparse+latent}$ should outperform the other estimators.

@@

Furthermore, $C_{\sf sparse}$ outperformed all the other estimators besides $C_{\sf sparse+latent}$ (Fig.~S\ref{supp:01}) suggesting the greater importance of modeling interactions between pairs of neurons than the common fluctuations of activity across the entire population.

@@

\paragraph{Structure of $C_{\sf sparse+latent}$ estimates}
Estimator $C_{\sf sparse+latent}$ separates two sources of correlations: a network of linear interactions between pairs of neurons and a set of latent units reflecting common fluctuations across the entire population.  These could have multiple physiological causes such as common inputs, cell assembly activation, network fluctuations, or other forms of synchronized activations.

We examined the composition of the $C_{\sf sparse+latent}$ estimates at each imaged site (Fig.~\ref{fig:05} and Fig.~\ref{fig:06}). Although the regularized estimates were similar to the sample correlation matrix (\figref{05}{A and D}), the corresponding partial correlation matrices differed substantially (\figref{05}{B and E}). The partial correlation matrix of the regularized estimate was the sum of a sparse component and a low-rank component (\figref{05}{C}).

In the example site (Fig.~\ref{fig:05}), the sparse component had 82.2\% sparsity (or conversely, 17.8\% connectivity: $connectivity=1-sparsity$) with average node degree of 52.9 (\figref{05}{G}). The averge node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $degree = connectivity\cdot(p-1)$. The low-rank component had rank 15, expressing 15 inferred latent units. The number of latent units and the average node degrees genearlly increased with population size, although these values varied widely between the datasets \figref{06}{A and B}. There was an inverse relationship between the number of latent units and the average node degree (\figref{06}{C}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and needs further investigation.

The average partial correlations calculated from these estimates according to Eq.~\ref{eq:partial} at all 31 sites were about 5 times lower than the average sample correlations (\figref{06}{D}). This signifies that correlations between neurons build up from large numbers of smaller partial correlations or interactions with other neurons. Furthermore, the avearge partial correlations were less variable: the coefficient of variation of the average sample correlations across sites was 0.41 whereas that of the average partial correlations was only 0.35.

While the sample correlations were mostly positive, the sparse component of the partial correlations (\sq{interactions}) had a high fraction (43.3\%) of negative (\sq{inhibitory}) values (\figref{05}{F}). This fraction was fairly consistent across multiple sites but the fraction of negative interactions increased with the inferred connectivity \figref{06}{F}.  This increase signifies that negative interactions are inferred only after a sufficient density of positive interactions have been uncovered.

@@

We compared the graphs of interactions inferred by $C_{\sf sparse+latent}$ to the graph obtained by thresholding the sample correlations.   Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013}.  We set the threshold to match the connectivity of the $C_{\sf sparse+latent}$ graph.  The networks revealed by the two methods differed substantially. In the example site with 17.8\% connectivity inferred by $C_{\sf sparse+latent}$, only 41.7\% of connections coincided with above-threshold sample correlations (\figref{05}{F, H, and I}). In particular, most of the inferred negative interactions corresponded to low sample correlations (\figref{05}{F}) where high correlations were expected given the rest of the correlation matrix.   This low overlap between the two types of graphs was consistent across across the datasets even though  their connectivity varied \figref{06}{E}.

@@

\paragraph{Relationship of $C_{\sf sparse+latent}$ to orientation tuning and physical distances}

We also examined the relationship of the correlation structure to the differences in orientation preference and to the physical distances between the cell pairs (Fig.\;\ref{fig:07}).  The partial correlations of the $C_{\sf sparse+latent}$ estimates fell more rapidly with the difference in preferred orientation (\figref{07}{A}) than the sample correlations. They also decreased more rapidly with physical distance, both laterally  (\figref{07}{B}) and in depth (\figref{07}{B}).

\begin{figure}[!ht]
    \begin{center}
        \includegraphics{./figures/Figure07.pdf}
    \end{center}
    \caption{{\bf Dependence of sample correlations and regularized partial correlations on orientation tuning and physical distance.}
    {\bf A--C.} Average partial correlations (red) estimated by $C_{\sf sparse+latent}$ and average sample correlations (black) averaged across multiple imaged sites. At each site the correlations were normalized by the respective average correlation shown in \figref{06}{D}.  The number of sites, $n$, that qualified to be included in the analysis is indicated in each panel. Sites were included if they had at least 20 pairs of neurons in each of the intervals. The error bars indicate the standard error of the mean based on on the number of sites $n$.
    {\bf A.} Average correlations between pairs of neurons tuned to orientation with differences in preferred orientation in the intervals of 0--15$^\circ$, 15--45$^\circ$ and 45--90$^\circ$. 
    {\bf B.} Average correlations between pairs of neurons located at the same depth ($\pm$25$\mu$m) separated by lateral distances in the intervals of 0--25 $\mu$m, 25--75 $\mu$m, 75--150 $\mu$m, and 150+ $\mu$m.
    {\bf C.} Average correlations between pairs of neurons displaced laterally by less than 25 $\mu$m separated in depth by distances in the intervals of 0--25 $\mu$m, 25--60 $\mu$m, and 60+ $\mu$m.
    {\bf D--F.} Normalized connectivity of positive (green) and negative (dark red) interactions from the sparse component obtained from $C_{\sf sparse+latent}$. Normalized connectivity was computed as the fraction of pairs connected by interactions of corresponding signs in each interval divided by the fraction of non-zero interactions across the entire site. }
\label{fig:07}
\end{figure}

The positive and negative interactions in the $C_{\sf sparse+latent}$ estimator were organized differently: The density of the positive interactions fell rapidly with differences in the preferred orientation (\figref{07}{D}), lateral displacements (\figref{07}{E}), and displacement in depth (\figref{07}{F}) whereas the negative interactions were much less selective (\figref{07}{D--F}).
